---
title: "Distributed vector quantization with barycentric prototypes"
author: "Kisung You"
format:
  gfm:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dpi: 400
---

## Overview

This notebook reproduces the distributed vector quantization (DVQ) experiment.
The goal is to evaluate whether free-support Wasserstein barycenters can be used
as compressed clustering prototypes.  The notebook compares DVQ with standard
`k`-means and spherical `k`-means.

The original manuscript experiment used PBMC, 20NEWS, and FASHION.  Because the
processed 20NEWS file too large to be uploaded to GitHub, this notebook is configured by default to run the reader-friendly two-dataset version using PBMC and FASHION only.  

For the default two-dataset run, place the following processed files in `../data/`:

- `processed_pbmc.RData`
- `processed_fmnist.RData`

Each file should contain a processed feature matrix `X_reduced` and a label vector
`Y`.

## Setup

Edit the first cell to choose a smoke test or the full reviewer-response grid.

```{r user-settings}
# User settings ------------------------------------------------------------
run_mode <- "smoke"       # "smoke" or "full"
run_compute <- TRUE       # FALSE: skip computation and only make figures from existing results
install_packages <- FALSE # TRUE: install missing R packages from CRAN when possible
jobs <- 1L                # use 1 for smoke; increase for full runs on a workstation
experiment_scale <- "core"
run_label <- ""           # automatic: "smoke-nonews" or "core-nonews"

# Data sets to include.  The default excludes 20NEWS because the processed
# file is large.  To reproduce the full three-dataset manuscript version, use
# c("pbmc", "news", "fashion") and put processed_news.RData in ../data/.
cluster_datasets <- c("pbmc", "fashion")
# -------------------------------------------------------------------------
```

```{r setup}
source("auxiliary.R")
load_or_install(
  c("T4transport", "skmeans", "cluster", "clusterCrit", "mclustcomp", "ggplot2", "patchwork", "dplyr", "readr", "tidyr"),
  install = install_packages
)
cluster_datasets <- match.arg(cluster_datasets, choices = c("pbmc", "news", "fashion"), several.ok = TRUE)
cluster_datasets <- unique(cluster_datasets)

# Use a distinct label for the default two-dataset run so that outputs are not
# mixed with older three-dataset results.
if (!nzchar(run_label) && !all(c("pbmc", "news", "fashion") %in% cluster_datasets)) {
  run_label <- if (run_mode == "smoke") "smoke-nonews" else "core-nonews"
}
run_label <- set_run_environment(run_mode, run_label, experiment_scale)

data_dir <- normalizePath(file.path("..", "data"), mustWork = TRUE)
required_data_files <- c(
  pbmc = "processed_pbmc.RData",
  news = "processed_news.RData",
  fashion = "processed_fmnist.RData"
)
for (dataset_name in cluster_datasets) {
  require_file(
    file.path(data_dir, required_data_files[[dataset_name]]),
    paste0(toupper(dataset_name), " processed data")
  )
}
Sys.setenv(
  CLUSTER_DATA_DIR = data_dir,
  CLUSTER_DATASETS = paste(cluster_datasets, collapse = ",")
)
message("Run mode: ", run_mode)
message("Run label: ", run_label)
message("Data sets: ", paste(cluster_datasets, collapse = ", "))
message("Using clustering data directory: ", data_dir)
```

## Computation

The full grid varies the selected data sets, cluster count, DVQ subset count, and
step size.  The smoke grid runs a single task to test data loading, clustering
metrics, saving, and assembly.

```{r compute-clustering}
if (run_compute) {
  if (run_mode == "smoke") {
    c1_tasks <- 1L
  } else {
    n_methods <- if (tolower(experiment_scale) == "heavy") 8L else 6L
    n_cluster_indices <- if (tolower(experiment_scale) == "heavy") 11L else 3L
    n_repetitions <- if (tolower(experiment_scale) == "heavy") 10L else 3L
    n_tasks <- length(cluster_datasets) * n_methods * n_cluster_indices * n_repetitions
    c1_tasks <- seq_len(n_tasks)
  }
  run_rscript_grid(
    file.path("real-2-clustering-extended", "01_clustering_alpha_runtime.R"),
    c1_tasks,
    jobs = jobs,
    label = "DVQ clustering"
  )
  run_rscript_once(file.path("real-2-clustering-extended", "02_assemble_clustering_alpha_runtime.R"))
}
```

## Figures and tables

In full mode, the visualization script produces the PCA visualization, clustering
quality figure, and summary tables for the selected data sets.  In smoke mode we
print the assembled result.

```{r visualize-clustering, fig.align="center", out.width="98%"}
if (run_mode == "full") {
  source("vis-real-clustering.R")
} else {
  cl_file <- file.path("real-2-clustering-extended", paste0("assembled_cluster_extended-", run_label, ".csv"))
  if (file.exists(cl_file)) print(readr::read_csv(cl_file, show_col_types = FALSE))
}
```

## Notes

The reference labels are used only for evaluation through ARI and NMI.  The
Silhouette and Calinski--Harabasz indices are internal geometric criteria.  This
default notebook omits 20NEWS only for repository-size reasons; the same code can
include it by changing `cluster_datasets` and adding the processed data file.

