---
title: "Bayesian posterior aggregation with Wasserstein barycenters"
author: "Kisung You"
format:
  gfm:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dpi: 400
---

## Overview

This notebook reproduces the Bayesian posterior aggregation experiment.  The
full-data posterior is available analytically for the simulated conjugate linear
model, while precomputed MCMC/subset posterior draws are supplied in the data
file.  The notebook then runs the free-support barycenter aggregation step and
visualizes Wasserstein, moment, predictive, coverage, and runtime diagnostics.

Before rendering this notebook, place the following file in `../data/`:

- `normal-simplified-reduced.RData`

## Setup

Edit only the first cell for a quick smoke run or a full reproduction.

```{r user-settings}
# User settings ------------------------------------------------------------
run_mode <- "smoke"       # "smoke" or "full"
run_compute <- TRUE       # FALSE: skip computation and only make figures from existing results
install_packages <- FALSE # TRUE: install missing R packages from CRAN when possible
jobs <- 1L                # use 1 for smoke; increase for full runs on a workstation
experiment_scale <- "core"
run_label <- ""           # automatic: "smoke" for smoke, "core" for full
# -------------------------------------------------------------------------
```

```{r setup}
source("auxiliary.R")
load_or_install(
  c("T4transport", "mvtnorm", "MASS", "ggplot2", "patchwork", "dplyr", "readr", "tidyr"),
  install = install_packages
)
run_label <- set_run_environment(run_mode, run_label, experiment_scale)
wasp_data <- require_file(file.path("..", "data", "normal-simplified-reduced.RData"), "WASP data file")
Sys.setenv(WASP_DATA_FILE = wasp_data)
message("Run mode: ", run_mode)
message("Run label: ", run_label)
message("Using WASP data: ", wasp_data)
```

## Computation

The full grid varies the number of subsets, barycenter support size, and step
size.  The smoke grid runs only a few tasks so that package loading, data access,
T4transport calls, and assembly can be checked quickly.

```{r compute-wasp}
if (run_compute) {
  w1_tasks <- if (run_mode == "smoke") 1:2 else 1:54
  run_rscript_grid(
    file.path("simulation-normal-extended", "01_wasp_alpha_barycenters.R"),
    w1_tasks,
    jobs = jobs,
    label = "WASP barycenters"
  )
  run_rscript_once(file.path("simulation-normal-extended", "02_assemble_wasp_alpha_metrics.R"))
}
```

## Figures and tables

The manuscript figures are generated in full mode.  Smoke mode prints the small
assembled result so the reader can confirm that the pipeline is working.

```{r visualize-wasp, fig.align="center", out.width="98%"}
if (run_mode == "full") {
  source("vis-simulation-normal.R")
} else {
  wasp_file <- file.path("simulation-normal-extended", paste0("assembled_wasp_alpha_metrics-", run_label, ".csv"))
  if (file.exists(wasp_file)) print(readr::read_csv(wasp_file, show_col_types = FALSE))
}
```

## Notes

The precomputed MCMC file is used so that the notebook focuses on the barycenter
aggregation step.  Regenerating all MCMC draws is much more expensive and is not
needed to check the barycenter computation.

