---
title: "MNIST digit prototypes and classification"
author: "Kisung You"
format:
  gfm:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dpi: 400
---

## Overview

This notebook reproduces the handwritten digit prototype experiment.  Images are
converted to point-cloud measures by Otsu thresholding, one free-support
Wasserstein barycenter is built for each class, and test images are classified by
nearest prototype in Wasserstein distance.  Euclidean class-mean and 1-nearest
neighbor baselines are included for reference.

The notebook uses the `digits` data bundled with `T4transport`; no external MNIST
file is needed.

## Setup

Edit the first cell to choose a smoke test or the full reviewer-response grid.

```{r user-settings}
# User settings ------------------------------------------------------------
run_mode <- "smoke"       # "smoke" or "full"
run_compute <- TRUE       # FALSE: skip computation and only make figures from existing results
install_packages <- FALSE # TRUE: install missing R packages from CRAN when possible
jobs <- 1L                # use 1 for smoke; increase for full runs on a workstation
experiment_scale <- "core"
run_label <- ""           # automatic: "smoke" for smoke, "core" for full
# -------------------------------------------------------------------------
```

```{r setup}
source("auxiliary.R")
load_or_install(
  c("T4transport", "ggplot2", "patchwork", "dplyr", "readr", "tidyr", "MASS"),
  install = install_packages
)
run_label <- set_run_environment(run_mode, run_label, experiment_scale)
message("Run mode: ", run_mode)
message("Run label: ", run_label)
```

## Computation

The full grid computes prototypes for all digits, support sizes, step sizes, and
three train/test splits.  The smoke grid computes enough prototypes for the first
classification task and then runs one baseline task.

```{r compute-digits}
if (run_compute) {
  if (run_mode == "smoke") {
    # Smoke classification task 1 corresponds to rep = 1, support = 40,
    # alpha = 1. Because the centroid task grid is generated by
    # expand.grid(rep, digit, support, alpha), the 10 needed class
    # prototypes have task ids 1, 4, 7, ..., 28, not 1:10.
    d1_tasks <- seq(from = 1L, to = 28L, by = 3L)
    d2_tasks <- 1L
    d3_tasks <- 1L
  } else {
    d1_tasks <- 1:180
    d2_tasks <- 1:18
    d3_tasks <- 1:3
  }

  run_rscript_grid(
    file.path("real-1-digits-extended", "01_digit_centroids_alpha.R"),
    d1_tasks,
    jobs = jobs,
    label = "Digit prototypes"
  )
  run_rscript_grid(
    file.path("real-1-digits-extended", "02_digit_classification_alpha.R"),
    d2_tasks,
    jobs = jobs,
    label = "Digit Wasserstein classification"
  )
  run_rscript_grid(
    file.path("real-1-digits-extended", "03_digit_euclidean_baselines.R"),
    d3_tasks,
    jobs = jobs,
    label = "Digit Euclidean baselines"
  )
  run_rscript_once(file.path("real-1-digits-extended", "04_assemble_digits.R"))
}
```

## Figures and tables

In full mode, the visualization script produces the preprocessing figure, digit
prototype figure, and summary tables.  In smoke mode we print the available
assembled metrics.

```{r visualize-digits, fig.align="center", out.width="98%"}
if (run_mode == "full") {
  source("vis-real-digits.R")
} else {
  files <- c(
    file.path("real-1-digits-extended", paste0("assembled_digit_centroids-", run_label, ".csv")),
    file.path("real-1-digits-extended", paste0("assembled_digit_classification-", run_label, ".csv")),
    file.path("real-1-digits-extended", paste0("assembled_digit_baselines-", run_label, ".csv"))
  )
  for (ff in files) if (file.exists(ff)) print(readr::read_csv(ff, show_col_types = FALSE))
}
```

## Notes

The digit experiment is best interpreted as a prototype-compression example, not
as a state-of-the-art classifier.  The Euclidean baselines are included to show
how much information is lost when grayscale images are reduced to thresholded
support geometry.

