---
title: "Gaussian barycenters: resolution, step size, and baselines"
author: "Kisung You"
format:
  gfm:
    toc: true
    toc-depth: 2
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
knitr:
  opts_chunk:
    dpi: 400
---

## Overview

This notebook reproduces the Gaussian barycenter experiment from the paper.  The
purpose is to study finite-resolution behavior in a setting where the population
barycenter is known, to compare the full update with a damped update, and to give
representative computational baselines using POT.

The notebook is self-contained except for software dependencies.  The Gaussian
inputs are generated synthetically, so no external data file is required.

## Setup

The first code cell is the only place where most users should edit the run.  Use
`run_mode <- "smoke"` for a quick check and `run_mode <- "full"` for the
manuscript-scale reviewer-response grid.  Full mode can be computationally
expensive because exact OT subproblems dominate runtime.

```{r user-settings}
# User settings ------------------------------------------------------------
run_mode <- "smoke"       # "smoke" or "full"
run_compute <- TRUE       # FALSE: skip computation and only make figures from existing results
install_packages <- FALSE # TRUE: install missing R packages from CRAN when possible
jobs <- 1L                # use 1 for smoke; increase for full runs on a workstation
experiment_scale <- "core"
run_label <- ""           # automatic: "smoke" for smoke, "core" for full
# -------------------------------------------------------------------------
```

```{r setup}
source("auxiliary.R")
load_or_install(
  c("T4transport", "mvtnorm", "abind", "ggplot2", "patchwork", "dplyr", "readr", "tidyr"),
  install = install_packages
)
run_label <- set_run_environment(run_mode, run_label, experiment_scale)
message("Run mode: ", run_mode)
message("Run label: ", run_label)
```

## Computation

The finite-resolution experiment uses support sizes and step sizes from the
paper.  The runtime experiment uses one-factor sweeps in support size, target
support size, number of measures, dimension, and step size.  The POT baseline
requires Python with the packages listed in `../requirements.txt`.

```{r compute-gaussian}
if (run_compute) {
  if (run_mode == "smoke") {
    # Enough to test data generation, T4transport calls, saving, and assembly.
    g1_tasks <- 1:2
    g2_tasks <- 1:2
    pot_tasks <- 1
  } else {
    # Core reviewer-response grid after reducing repetitions to keep exact OT feasible.
    g1_tasks <- 1:36
    g2_tasks <- 1:54
    pot_tasks <- 1:27
  }

  run_rscript_grid(
    file.path("simulation-gaussian-extended", "01_gaussian_resolution_alpha.R"),
    g1_tasks,
    jobs = jobs,
    label = "Gaussian resolution / alpha"
  )
  run_rscript_once(file.path("simulation-gaussian-extended", "02_assemble_gaussian_resolution_alpha.R"))

  run_rscript_grid(
    file.path("simulation-gaussian-extended", "03_runtime_scaling.R"),
    g2_tasks,
    jobs = jobs,
    label = "Gaussian runtime scaling"
  )
  run_rscript_once(file.path("simulation-gaussian-extended", "04_assemble_runtime_scaling.R"))

  # POT is optional in smoke mode if Python/POT is unavailable. In full mode,
  # it is part of the baseline comparison used in the paper.
  try({
    run_python_grid(
      file.path("pot-baselines", "01_pot_gaussian_baselines.py"),
      pot_tasks,
      jobs = jobs,
      label = "POT baselines"
    )
    run_python_grid(
      file.path("pot-baselines", "02_assemble_pot_gaussian_baselines.py"),
      1,
      jobs = 1,
      label = "POT assembly"
    )
  }, silent = (run_mode == "smoke"))
}
```

## Figures and tables

In full mode, the following visualization script reproduces the manuscript
figures and summary tables.  In smoke mode, the same script may have too little
data for all panels, so we print the assembled smoke summaries instead.

```{r visualize-gaussian, fig.align="center", out.width="98%"}
if (run_mode == "full") {
  source("vis-simulation-gaussian.R")
} else {
  g1_file <- file.path("simulation-gaussian-extended", paste0("assembled_gaussian_resolution_alpha-", run_label, ".csv"))
  g2_file <- file.path("simulation-gaussian-extended", paste0("assembled_runtime_scaling-", run_label, ".csv"))
  if (file.exists(g1_file)) print(readr::read_csv(g1_file, show_col_types = FALSE))
  if (file.exists(g2_file)) print(readr::read_csv(g2_file, show_col_types = FALSE))
}
```

## Notes

The full update `alpha = 1` is the default because it exactly minimizes the
frozen-plan quadratic surrogate when OT subproblems are solved exactly.  The
damped update `alpha = 0.5` is included as a sensitivity analysis.

